---
title: "Assignment 11: DNA methylation and enrichment analysis"
author: "Lukas Häuser"
output:
  html_document: default
  pdf_document: default
---

```{r}
set.seed(40) # Fix seed for reproducibility
```

Get Bisulfite-sequencing data:
```{r, bsseq}
bs <- readRDS("./w11_practical/bs.rds")

MatrixGenerics::rowRanges(bs)
Biobase::pData(bs)
```
Get annotations (hs):
```{r}
# genes
ah <- AnnotationHub::AnnotationHub()
ensdb <- ah[["AH109336"]]
chr22 <-  GRanges(seqnames=Rle(c("22")), 
                  ranges = IRanges(1, end=50818468))
genesChr22 <- genes(ensdb, columns=c("gene_seq_start", "gene_seq_end", "gene_name"),
                    filter=GRangesFilter(chr22))
seqlevelsStyle(genesChr22) <- "UCSC"
```

Obtain the coordinates of the genes within the top 5 DMRs.  
```{r}
dmrRanges <- readRDS("./w11_practical/dmr.rds")

dmrRanges <- dmrRanges[order(abs(dmrRanges$meandiff), decreasing=TRUE)]
dmrRangesGenes <- dmrRanges[!is.na(dmrRanges$overlapping.genes)]

# Get the genes within Differentially methylated regions
topIdx <- order(dmrRangesGenes$min_smoothed_fdr)[1:5] # get top 5 DMR regions
genesDmr <- unlist(data.table::tstrsplit(dmrRangesGenes[topIdx]$overlapping.genes, split=", "))
genesDmr <- genesDmr[!is.na(genesDmr)]
dmrGenes.top <- genesChr22[genesChr22$gene_name %in% genesDmr]
dmrGenes.top

# Get all
genesDmr <- unlist(data.table::tstrsplit(dmrRangesGenes$overlapping.genes, split=", "))
genesDmr <- genesDmr[!is.na(genesDmr)]
dmrGenes.all <- genesChr22[genesChr22$gene_name %in% genesDmr]
dmrGenes.all
```
Retrieve metyhlation levels and visualize:
```{r, look at bsseq objects}
metPr <- bsseq::getMeth(bs, 
                        regions=dmrGenes.top, 
                        what="perRegion")
colnames(metPr) <- colnames(bs)
rownames(metPr) <- dmrGenes.top$gene_name
metPr <- metPr[!is.na(rowSums(metPr)),]

annotationCol <- as.data.frame(pData(bs))
rownames(annotationCol) <- colnames(metPr)
pheatmap::pheatmap(metPr, 
                   cluster_rows=TRUE,
                   cluster_cols=FALSE,
                   annotation_col=annotationCol,
                   show_rownames = TRUE,
                   color=viridis::rocket(10))
```
Enrichment analysis of all genes within DMRs agaist all genes on chr22:
```{r}
res <- rGREAT::great(dmrGenes.all, gene_sets="GO:BP", tss_source="hg38", background=genesChr22, cores=6)
bp <- rGREAT::getEnrichmentTables(res)
bp.significant <- bp[bp$p_adjust < 0.05,] # filter for significance with p<5%
bp.significant
```

Enrichment analysis of genes within top5 DMRs agaist all genes on chr22 (just because we thought that might be interesting too):
```{r}
res <- rGREAT::great(dmrGenes.top, gene_sets="GO:BP", tss_source="hg38", background=genesChr22, cores=6)
bp <- rGREAT::getEnrichmentTables(res)
bp.significant <- bp[bp$p_adjust < 0.05,] # filter for significance with p<5%
bp.significant
```

# Results

We identified the 5 regions with highest differential methylation. These 5 regions contain 7 genes (CDC45, PARVG, SHISAL1, FBLN1, PKDREJ, TRMU, CELSR1). We did enrichment analyses of all genes within DRMs against all genes of chr22 and of the 7 genes within the top5 DRMS. Then we retrieved the methylation levels of these 7 genes. With a significance of `p_adust<5%`, we found 68 enriched gene sets within DRMs on chr22. The highest fold enrichment of these gene sets is about 5.2 between the normal and cancer cases. In the the genes within the top 5 DRMs only a single gene set was found with a particularly high fold enrichment of approximately 7.2. Thus, it seems like the cancer case could be associated with high fold enrichment of highly diffirentially methylated regions.